\subsection*{Representation}
The traditional approach to flocking simulation maintains a matrix
representing the simulated universe which contains birds at specified
indices. This representation is idea for graphics applications, since
the state of the universe is explicitly defined. However, the explicit
representation is less than ideal for scientific reasons--- while
flocking behavior can be simulated correctly and efficiently, the
manner in which the behavior is simulated provides little predictive
power with regard to the cognitive, perceptual, and social mechanisms
required to exhibit flocking behavior. Additionally the explicit
representation has significant memory limitations. Since the entire
state of the universe must be stored, the universe may only be as
large as the total amount of memory available.

In this project we take an agent-based perspective, which favors
computation from the perspective of the individual rather than the
system as a whole. This representation is more cognitively plausible,
since computation must generally follow the steps of each individual
agent. Using the agent-based approach, the universe is represented
\emph{implicitly}. Individual locations are stored locally by the
agent rather than by the individual's index within a global matrix,
meaning that an increase in universe size does not incur additional
memory costs. 

\subsection*{Serial Algorithm}
The standard serial flocking simulation follows a familiar simulation
format. Roughly, calculate the three rules for each bird, then apply
the rules to each bird after all calculations are complete.

\newpage
\begin{verbatim}
  for ( time-steps 0...t ) {
    for ( Bird b ) {
      N = b's flock neighbors
      alignment = avg. direction over N
      cohesion = avg. x,y,z pos. over N
      separation = -avg. distance to n in N

      calculate dx/dy/dz
    }

    for ( Bird b )
     update b's position
  }
\end{verbatim}

\subsection*{Parallel Algorithm}
The parallel implementation of the simulation operates in much the
same way, with the provision that each process (MPI rank) is
responsible for a subset of birds of the flock. Processes must then
communicate their local information to all other processes in order
for update rules to include all applicable birds. This communication
is implemented using MPI\_Allgather; each rank recieves a complete
array of all birds in the universe to use in computation, but is only
responsible for calculating the position updates for its own agents.
Because MPI\_Allgather is a collective operation, ranks are
synchronized over each time-step, and therefore the simulation is
cycle-accurate.

We extend this approach to include functionality for
multithreading. Where each rank is assigned to a subset of agents,
each constituent thread of a process is responsible for a subset of
its rank's agents. MPI communication is only performed by the parent
thread, and threads are synchronized between time-steps.

This leads us to another benefit of an implicit representation---
where an explicit representation would be parallelized by assigning
each rank to a sector of the universe, the implicit representation
assigns agents to ranks. Computation is then more evenly distributed
between processes with the implicit representation, since with the
explicit representation entire sectors of the universe could be empty.
